State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-09-24 analysis/paper/paper.Rmd
removed comments that have been resolved.
Deleted a reference to a manuscript in review. re-knit

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

All captured fish were anesthetized, speciated, examined for existing marks or tags, measured for fork length and visually identified as wild or hatchery. The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are used in less than 10% of the hatchery releases. Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips (HNC). For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken. Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the location of origin of wild fish using genetic stock identification (e.g., @Hargrove2019). Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics. Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild. @Camacho2018 provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tagging [@Steele2013; @Steele2019]. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2018].

@@

<!-- John Hargrove: Is there a reason why you only present a singular year in Figure 3? I acknowledge that having 10 years of data may make it unreadable – but perhaps justify why 2014 was selected or include all 10 years as separate plots using facet_wrap in ggplot.

I think showing 10 time series plots is unnecessary (Table 3 shows how estimates match up with window counts). I think 2014 is a nice illustration of general agreement, with some examples of how STADEM doesn't always track the trap estimate, and how it deals with missing data. Do we need to make that more clear, or leave it as is?

After experimenting with 4 years, Ryan and Kevin decided even 4 was too busy. Let's stick with one, and note that the rest are on the github repo-->

@@

STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers. Currently, the only roadblock to this at Lower Granite Dam is the identification of hatchery no-clip fish using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season. The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement [@Hargrove2020]. However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement. Final post-hoc estimates parsed by origin could then be finalized at season’s end. All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in near real-time by DART. Provided the Lower Granite Dam adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.

@@

<!-- John Hargrove: Have you considered comparing estimate that you produced  to those that failed to account for fallback/reascension to see how much difference there is? In other words, you could draw estimates from Camacho and Schrader docs and readily compare – may serve to better illustrate your point of the superiority of accounting for these errors relative to less complicated models

I quickly pulled together the estimates from IDFG's latest report, 2010-2018 wild escapement. Their estimates of wild Chinook are about 4% higher than STADEM, and about 3% lower for wild steelhead. So far we've avoided explicitly comparing our results with existing methods in an effort to avoid confrontation. Maybe that's moot if IDFG has started adopting STADEM results, but looks like as of March 2019 they had not.

Ryan and Kevin decided to not go down this path. One quick comparison is total escapement vs. window counts; which shows up in one of our tables, and we discuss it briefly in the results. The other is figure 4, showing nighttime passage and re-ascension rates, and how they don't match. -->

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

All captured fish were anesthetized, speciated, examined for existing marks or tags, measured for fork length and visually identified as wild or hatchery. The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are used in less than 10% of the hatchery releases. Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips (HNC). For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken. Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the location of origin of wild fish using genetic stock identification [e.g., @Hargrove2019]. Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics. Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild. @Camacho2018 provide further details on trap sample rates and valid sample selection. Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tagging [@Steele2013; @Steele2019]. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2018].

@@

@@

STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers. Currently, the only roadblock to this at Lower Granite Dam is the identification of hatchery no-clip fish using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season. The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement. However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement. Final post-hoc estimates parsed by origin could then be finalized at season’s end. All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in near real-time by DART. Provided the Lower Granite Dam adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.

@@
